Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-08-25 doc/papers/cgo2014/paper.tex
Running example section

\subsection{Temporaries Elemination}

@@

\subsection{Loop Vectorization}

\subsection{Loop Parallelization}

>>>>

Let us consider the kernel illustrated in Listing~\ref{lst:rosen} and adapated
from the Scipy source code for \texttt{scipy.optimize.rosen}. This kernel makes
use of Numpy's \texttt{sum} function, Python's square notation and Numpy's
array slicing. It is a good example of high-level Python kernel, altohough
using the function \texttt{scipy.optimize.rosen} would naturally make sense.

\begin{lstlisting}[language=python, caption={High-level implementation of the rosenbrock function in Numpy}, float, label={lst:rosen}]
def rosen(x):
    t0 = 100 * (x[1:] - x[:-1] ** 2) ** 2
    t1 = (1 - x[:-1]) ** 2
    return numpy.sum(t0 + t1)
\end{lstlisting}

This section goes through all the optimization opportunities in
Listing~\ref{lst:rosen} as a showcase of what an optimizing compiler could do.

\subsection{Temporaries Elimination}

In Numpy, all point-to-point array operations allocate a new array that holds
the computation result. This behavior is consistent with many Python standard
module, but it is a very inefficient design choice, as it keeps on polluting
the cache with potentially large fresh storage. In the rosen expression from
Listing~\ref{lst:rosen}, 7 temporary arrays are allocated (slicing does not
create a temporary array but a view) to hold intermidiate steps. had the
computation been lazy, then no temporary would have been needed.

@@

As Numpy is a native library mostly written in C, each operator computation is
performed by a function is implemented as a loop performing a single operation,
while the operator chaining is done at the interpreter level. This a typical
problem in library design: if only a small set of functions is provided, it
prevents the optimization of multiple operators into a signle specialized
operator, and provided many operator combinations as part of the library yields
better performance to the price of API bloat. In that case, a loop is used for
each temporary computation, plus an extra loop for the \texttt{numpy.sum}
reduction, where a single loop would have been needed.

\subsection{Loop Vectorization and Parallelization}

Even if compiled with vectorization flags on, there would be very little
benefit to generate SIMD instructions for the respective array operations used
by each operator, as the memory load store would have dominated, especially as
Numpy typically operates on double precision floats. In a similar manner,
parallelization of each operator would yield terrible speedups because a
synchronization fence would need to be issued between each temporary
computation and the loop computation intensity would be very low compared to
the memory pressure implied by two array reads and one array write for a single
binary operator.

On the opposite, if all computation were merged into a single loop, the whole
loop would be triially vectorizable and parallelizable. No dependency checks
are needed as parallelization and vectorization are deduced from the semantics
of the array operations.
